<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Mutable Vectors in R}
-->

```{r global_options, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE)
```


# Mutable Vectors in R

R vectors are conceptually immutable. Replacement functions create a
modified copy of the original vector. So in this example the value of
`y` is unchanged:

```{r}
x <- 1:3
y <- x
x[1] <- 7
y
```
```{r, include = FALSE}
stopifnot(y[1] == 1)
```

This computational model avoids many bugs that might occur with a pass
by reference model, but it comes at the price of the overhead of
duplicating objects before modifying them. R does try to avoid
duplication when it can prove doing so is safe, but sometimes it
cannot be voided.

To avoid this potential overhead when initially building up an object,
or to maintain state, it might be useful to have R vectors that are
mutable, or have reference semantics, but otherwise behave like
ordinary R vectors. S3 or S4 classes can be used to some extent to
achieve this, but the resulting objects will not behave like ordinary
R vectors in all respects, in particular at the `C` level.

The `mutable` package was created to explore whether the `ALTREP`
framework can be used to create mutable R vectors that behave
properly. The answer seems to be no, at least at this point.


## The `mutable` Package

The `mutable` package provides an attempt to implement of mutable
vectors in R using the `ALTREP` framework.

Mutable vectors are created by `mutable`:
```{r}
library(mutable)  
(x <- mutable("double", 2))
y <- x
x[1] <- 1
y
```
```{r, include = FALSE}
stopifnot(x[1] == y[1]) 
```

The data portions of `x` and `y` are shared, so modifying the data of
`x` also modifies the data of `y`.  Only the data part of the vector
is shared; attribute assignments obey standard R semantics:

```{r}
attr(x, "foo") <- 1
attributes(y)
```
```{r, include = FALSE}
stopifnot(is.null(attributes(y)))
```

The function `is.mutable` tests whether its argument was created by
`mutable` (it does not allow for wrappers at this point).

```{r}
is.mutable(x)
is.mutable(y)
is.mutable(1:2)
```
```{r, include = FALSE}
stopifnot(is.mutable(x), is.mutable(y), ! is.mutable(1:2))
```

Possible additions, that could be done at the R level:

- Allow specifying initial data in the `mutable()` call.
- Add a `freeze` or `unmutable` function.


For now, only `integer` and `double` vectors are supported. Other
atomic types shoud be easy to add. Support for `character` vectors is
also probably straight forward. Vectors of type `list` may be more
challenging as cycles need to be prevented.

There is no special serialization support: A mutable vector is
serialized as a standard unmutable vector. Sharing could be preserved
within a serialization by using an external pointer as an
intermediary.  It's not clear this would be worth while.

There is no support for growing or shrinking a mutable vector.
Support along the lines of Common Lisp fill pointers could be
added. It's not clear whether integrating this with the growable
vector concept used when assigning beyond the current length of a
vector sould be a good idea.

At the `C` level both deep and shallow duplicating preserves data
sharing. Internally the object uses a separate R vector for the shared
storage, but this vector should not be accessed outside the package
`C` code.


## Problems

The base R code is written for the standard immutable vector
model. This leads to certain coding practices that are not compatible
with allowing vectors to have mutable data. Most of these are based on
the assumption that an object with `NO_REFERENCES` can be modified in
place r re-used. Two particular practices are:

- It is common at the `C` level to `duplicate` or `shallow_duplicate`
  an object and then modify it.

- In arithmetic operations in particular an argument with no
  references may be re-used for the result to avoid allocating a
  result vector.

### Duplicate and Modify

The duplicate and modify approach is This approach is based on the
assumption that `duplicate` and `shallow_duplicate` vector that have
no references and can be freely modified.  This approach is used in
many unary operations, including unary minus:

```{r}
x <- mutable("double", 2); x[] <- 1:2
-x
x
```

The attempt to create a mutable vector by having the `Duplicate`
method share the data portion violates the `duplicate` and
`shallow_duplicate` and would require modifying base R code, and
probably many packages, in many places to make it work properly.

Addressing this issue would require every use of the duplicate and
modify idiom to be replaced by allocating a new standard vector and
copying data and attributes as needed. The effort would be
significant. It would also loose the ability to automatically reuse
`ALTREP` objects that is currently possible, unless we added a
`Can_reuse` method to determine whether this is OK or not.
	     

### Value Re-Use

Without re-use an arithmetic operation with a mutable vector would
produce a standard vector result:

```{r}
x <- mutable("double", 2)
is.mutable(x + 1)
```

But when the mutable vector has no references it maybe re-used as the result:

```{r}
is.mutable(mutable("double", 2) + 1)
```

Reusing an unreferenced mutable vector as a result can allow
computations to unexpectedly produce mutable results, thus risking the
bugs that the basic R design is intended to prevent.

Marking the result produced by `mutable` as having references could
avoid this issue but not the duplicate-and-modify issue. It would also
force duplication of the `ALTREP` cell for any update operation.


<!--
Local Variables:
mode: poly-markdown+R
mode: flyspell
End:
-->
